Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-05 TOMS_Paper/assembly.tex
Rewrite Rewrite rules. General clean up

Consider the variant of the running (transformed) example shown in Figure~\ref{fig:toexpand-code}. Again, this is a representative example of what happens in real finite element forms. We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of \texttt{T1}, resulting in the code illustrated in Figure~\ref{fig:expanded-1-code}. Freeing the register destined to the constant \texttt{g} is more complicated: we cannot absorb it in the pre-computation of \texttt{T1} because the same array is accessed in the evaluation of \texttt{(T1[ j]*A[i][k])}. The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}. Generalizing, this is a problem of data dependencies; in order to solve it, we employ a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}. The dependency graph is initially empty; every time a new temporary is created due to loop-invariant code motion or expansion of terms is performed, it is updated by suitably adding vertices and edges.

@@

To address these issues, we have based the implementation of the ER in COFFEE on a set of formal rewrite rules. By applying these rules, it is possible to derive how an expression will be transformed, as well as what and where (i.e. at which level in the loop nest) temporaries will be introduced. When applying a rule, the ER needs to update the state of the loop nest, to reflect, for example, the use of a new temporary and the newly created data dependencies. We define, therefore, the state of a loop nest $L = (\sigma, G)$, where $G = (V, E)$ represents the dependency graph, while $\sigma$ maps invariant sub-expressions to identifiers of temporary arrays. We also introduce the \textit{conditional hoister} operator $[]$ on $\sigma : Inv \rightarrow S$ such that
\begin{gather*}
\sigma[\sfrac{v}{x}] = 
\begin{cases}
\sigma(x) \text{~~~~~~~~if $x \in Inv$; $v$ is ignored}\\
v \text{~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $\sigma$(x) = v}\\
\end{cases}
\end{gather*}
That is, intuitively, if the invariant expression \texttt{x} has already been hoisted, then return the temporary identifiers that hosts its value; otherwise, hoist the expression. There is a special case when $v = \perp$, used to delete entries in $\sigma$. Specifically:
\begin{gather*}
\sigma[\sfrac{\perp}{x}] = 
\begin{cases}
\sigma(x) \text{~~~~~~~~if $x \in Inv$; $\sigma = \sigma \setminus (x, \sigma(x))$}\\
t \text{~~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $t \notin Inv$}\\
\end{cases}
\end{gather*}
In other words, the previously hoisted expression \texttt{x} is removed (if any) and the temporary identifier that was hosting its value is returned. This is useful to express updates of invariant expressions.
Rewrite rules for the ER are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity. Conceptually, the ER visits the expression tree from the root, which is the outermost operation, and applies the transformations dictated by the rewrite rules. As an example, one can try instantiating the rules in the code of Figures~\ref{fig:original-code} and~\ref{fig:toexpand-code}; eventually, the optimized code in Figures~\ref{fig:factorized-code} and~\ref{fig:expanded-2-code} is obtained, respectively.

\begin{figure}
\small
\centering
\begin{spacing}{1.5}
\begin{align*}
[a_i \cdot b_j]_{(\sigma, G)} &\rightarrow [a_i \cdot b_j]_{(\sigma, G)} ~~&~~&\\
[(a_i + b_j)\cdot \alpha]_{(\sigma, G)} &\rightarrow [(a_i \cdot \alpha + b_j \cdot \alpha)]_{(\sigma, G)} ~~&~~ &\\
[a_i \cdot b_j + a_i \cdot c_j]_{(\sigma, G)} &\rightarrow [(a_i \cdot (b_j + c_j)]_{(\sigma, G)} ~~&~~ &\\
[a_i + b_i]_{(\sigma, G)} &\rightarrow [t_i]_{(\sigma', G')} ~~&~~ &t_i = \sigma[\sfrac{t_i'}{a_i + b_i}], G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, b_i)\rbrace)\\
[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma[\sfrac{\sigma[\sfrac{\perp}{a_i}]}{a_i \cdot \alpha}], a_i \notin in(G), \\
~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma[\sfrac{t_i'}{a_i \cdot \alpha}], a_i \in in(G), \\
~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
\end{align*}
\end{spacing}
\caption{Rewrite rules.}\label{fig:rewrite-rules}
\end{figure}

@@

%\begin{figure}[t]
%\tiny
%\centering     
%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p1_formhelmholtz_speedup}}
%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p2_formhelmholtz_speedup}}
%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p3_formhelmholtz_speedup}}
%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p4_formhelmholtz_speedup}}
%\caption{Helmholtz results.}\label{fig:helmholtz}
%\end{figure}

@@

>>>>

Consider the variant of the running (transformed) example shown in Figure~\ref{fig:toexpand-code}. Again, this is a representative example of what happens in real finite element forms. We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of \texttt{T1}, resulting in the code illustrated in Figure~\ref{fig:expanded-1-code}. Freeing the register destined to the constant \texttt{g} is more complicated: we cannot absorb it in the pre-computation of \texttt{T1} because the same array is accessed in the evaluation of \texttt{(T1[j]*A[i][k])}. The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}. Generalizing, this is a problem of data dependencies; in order to solve it, we employ a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}. The dependency graph is initially empty; every time a new temporary is created due to loop-invariant code motion or expansion of terms is performed, it is updated by suitably adding vertices and edges.

@@

To address these issues, we have based the implementation of the ER in COFFEE on a set of formal rewrite rules. By applying these rules, it is possible to derive how an expression will be transformed, as well as what and where (i.e. at which level in the loop nest) temporaries will be introduced. When applying a rule, the ER needs to update the state of the loop nest, to reflect, for example, the use of a new temporary and the newly created data dependencies. We define, therefore, the state of a loop nest $L = (\sigma, G)$, where $G = (V, E)$ represents the dependency graph, while $\sigma$ maps invariant sub-expressions to identifiers of temporary arrays. The notation $\sigma_i$ refers to invariants hoisted at the level of loop $i$. We also introduce the \textit{conditional hoister} operator $[]$ on $\sigma : Inv \rightarrow S$ such that
\begin{gather*}
\sigma[\sfrac{v}{x}] = 
\begin{cases}
\sigma(x) \text{~~~~~~~~if $x \in Inv$; $v$ is ignored}\\
v \text{~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $\sigma$(x) = v}\\
\end{cases}
\end{gather*}
That is, intuitively, if the invariant expression \texttt{x} has already been hoisted, then return the temporary identifiers that hosts its value; otherwise, hoist the expression. There is a special case when $v = \perp$, used to delete entries in $\sigma$. Specifically:
\begin{gather*}
\sigma[\sfrac{\perp}{x}] = 
\begin{cases}
\sigma(x) \text{~~~~~~~~if $x \in Inv$; $\sigma = \sigma \setminus (x, \sigma(x))$}\\
t \text{~~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $t \notin Inv$}\\
\end{cases}
\end{gather*}
In other words, the previously hoisted expression \texttt{x} is removed (if any) and the temporary identifier that was hosting its value is returned. This is useful to express updates of invariant expressions.

A generic (sub-)expression is represented with roman letter \texttt{a, b, ...}; constant terms are considered a special case, so greek letters \texttt{$\alpha$, $\beta$} are used instead. The iteration vector $i = [i_0, i_1, ...]$ is the ordered sequence of the indices of the loops enclosing an (sub-)expression. We will refer to $i_0$ as the outermost enclosing loop. The notation $a_i$, therefore, indicates that the expression $a$ assume distinct values while iterating along the loops in $i$; its outermost loop is silently assumed to be $i_0$.

Rewrite rules for the ER are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity. Conceptually, the ER visits the expression tree from the root, which is the outermost operation, and applies the transformations dictated by the rewrite rules. As an example, one can try instantiating the rules in the code of Figures~\ref{fig:original-code} and~\ref{fig:toexpand-code}; eventually, the optimized code in Figures~\ref{fig:factorized-code} and~\ref{fig:expanded-2-code} is obtained, respectively.

\begin{figure}
\small
\centering
\begin{spacing}{1.35}
\begin{align*}
~~~~~~&Rule &~~&~~~~~~~~~~~~~~~~~~~~~~~~Precondition \\
[a_i \cdot b_j]_{(\sigma, G)} &\rightarrow [a_i \cdot b_j]_{(\sigma, G)} ~~&~~&\\
[(a_i + b_j)\cdot \alpha]_{(\sigma, G)} &\rightarrow [(a_i \cdot \alpha + b_j \cdot \alpha)]_{(\sigma, G)} ~~&~~ &\\
[a_i \cdot b_j + a_i \cdot c_j]_{(\sigma, G)} &\rightarrow [(a_i \cdot (b_j + c_j)]_{(\sigma, G)} ~~&~~ &\\
[a_i + b_i]_{(\sigma, G)} &\rightarrow [t_i]_{(\sigma', G')} ~~&~~ &t_i = \sigma_{i_0}[\sfrac{t_i'}{a_i + b_i}], G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, b_i)\rbrace)\\
[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma_{i_0}[\sfrac{\sigma[\sfrac{\perp}{a_i}]}{a_i \cdot \alpha}], a_i \notin in(G), \\
~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma_{i_0}[\sfrac{t_i'}{a_i \cdot \alpha}], a_i \in in(G), \\
~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
\end{align*}
\end{spacing}
\caption{Rewrite rules.}\label{fig:rewrite-rules}
\end{figure}

@@

\begin{figure}[t]
\centering     
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p1_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p2_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p3_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p4_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p1_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p2_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p3_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p4_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p1_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p2_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p3_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p4_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p1_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p2_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p3_formhelmholtz_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p4_formhelmholtz_speedup}}
\caption{Helmholtz results.}\label{fig:helmholtz}
\end{figure}

In Figure~\ref{fig:elasticity}, ...

\begin{figure}[t]
\centering     
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p1_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p2_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p3_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p4_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p1_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p2_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p3_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p4_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p1_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p2_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p3_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p4_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p1_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p2_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p3_formelasticity_speedup}}
\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p4_formelasticity_speedup}}
\caption{Elasticity results.}\label{fig:elasticity}
\end{figure}

@@

We have presented ...
